---
title: "Sentiment Analysis in Crude Oil Markets"
date: Last updated as of `r Sys.Date()`
output: 
  github_document:
    toc: true
    toc_depth: 3
---

## Project Summary

This project leverages on text data sourced from [Twitter](https://twitter.com/) in addition to Natural Language Processing packages in R (E.g. [qdap](https://www.rdocumentation.org/packages/qdap/versions/2.3.2) and [tidytext](https://www.rdocumentation.org/packages/tidytext/versions/0.2.0)) to perform basic Sentiment Analysis.

The objective will be to see if applying such methods to Twitter-sourced text data leads to any interesting insights. Future implementations might involve the integration of such findings with an algorithmic trading strategy to take positions based on market sentiment.

## Package Imports

```{r message=FALSE}
library(knitr)  # Export R Notebook to R Script
library(qdap)  # Quantitative Discourse Analysis Package
library(RCurl)  # For HTTP Requests
library(stringr)  # For String Manipulation
library(tidytext)  # Natural Language Processing package in R
library(tidyverse)  # For data wrangling
library(tm)  # Text Mining
library(wordcloud)  # Generation of Word Clouds
```

## Code

### 1. Import Text Data

```{r read_data, message=FALSE}
crude_df <- read_csv("tweets.csv")
```

### 2. Text Preprocessing

```{r preprocess_text}
# Remove all preceding "RT" characters in Retweeted tweets
crude_df$text <- gsub(pattern = "RT", replacement = "", crude_df$text)

# Remove all Twitter user handles in Retweets
crude_df$text <- gsub(pattern = "@\\w+", replacement = "", crude_df$text)

# Convert all tweets to lower case
crude_df$text <- tolower(crude_df$text)

# Remove punctuation
crude_df$text <- removePunctuation(crude_df$text)

# Remove quotation marks
crude_df$text <- gsub(pattern = '“', replacement = "", crude_df$text)

# Remove URL links
crude_df$text <- gsub("http\\w+", "", crude_df$text)

# Replace contractions
crude_df$text <- replace_contraction(crude_df$text)

# Define a list of uninformative stop-words for removal
crude_stopwords <- c("crudeoil", "Crudeoil", "crude", "oil", "oott", stopwords("en"))

# Remove all stop words
crude_df$text <- removeWords(crude_df$text, crude_stopwords)

# Remove all whitespace
crude_df$text <- stripWhitespace(crude_df$text)

# Examine the text data without stop words
crude_df$text[1:10]
```

The Twitter text data at this point appears to be sufficiently pre-processed. Next steps in the project will include basic analytics with wordclouds and ggplot, before moving onto Sentiment Analysis.

### 3. Exploratory Data Analysis with ggplot and wordcloud

```{r plot_wordfreqs}
# Explore the most common words in the data set
crude_df %>%
  unnest_tokens(word, text) %>%
  group_by(word) %>%
  count() %>%
  arrange(desc(n)) %>%
  head(n = 10) %>%
  ggplot(mapping = aes(x = reorder(word, desc(n)), y = n, fill = word)) +
    geom_col() +
    labs(
      title = "Word Frequency in #crudeoil tweets",
      subtitle = paste("Data collected on", Sys.Date()),
      caption = "Source: Twitter API",
      x = "Word",
      y = "Frequency") +
    theme_minimal() +
    theme(legend.position = "none")
```


```{r plot_wordcloud, warning=FALSE}
# Generate a simple word cloud on the Twitter data
wordcloud(
  words = crude_df$text, 
  min.freq = 15, 
  colors = brewer.pal(8, "Dark2"), 
  random.color = TRUE, 
  max.words = 250
  )
```

### 4. Basic Sentiment Analysis with qdap::polarity()

Next, apart from simple word counts, we want to roughly estimate the "polarity" or emotion of the collection of tweets. To do this, we use the polarity() function from the qdap package.

```{r compute_polarity}
# Define a Polarity Object for the set of #crudeoil tweets
crude_polarity <- polarity(crude_df$text)

# Print a summary of the polarity object
summary(crude_polarity$all$polarity)
```

```{r plot_polarity, warning=FALSE}
# Visualize the polarity object with ggplot
ggplot(crude_polarity$all, aes(x = polarity, y = ..density..)) +
  geom_histogram(binwidth = 0.25, fill = "#bada55", colour = "grey60") +
  geom_density(color = "darkblue") +
  labs(
    title = "qdap Polarity Scores of #crudeoil tweets",
    subtitle = paste("As of", Sys.Date()),
    caption = "Source: Twitter API",
    x = "Polarity Score",
    y = "Frequency"
  )
```

A rough analysis of all tweets in the dataset does not suggest that sentiment in this space skews either left or right. Most observations appear to return a polarity score of 0, although the mean does appear to tilt to the left (i.e. negativity).

Further analysis on sentiment will be attempted by joining a pre-loaded Lexicon of sentiment and repeating the analysis.

### 5. Preprocessing for tidytext

Another method of drawing sentiment from textual data involves using the tidytext package and its associated lexicons (E.g. NRC, AFINN and Bing). In this following section, we perform an inner-join of our crude data to the lexicons and attempt to draw further insights.

```{r store_vector}
# Store a character vector containing all tweets
tweet_vector <- crude_df$text

# Examine the vector
head(tweet_vector)
```

```{r create_corpus}
# Convert the character vector into a Volatilte Corpus
tweet_corpus <- VCorpus(VectorSource(tweet_vector))

# Convert the corpus into a Document Term Matrix
tweet_dtm <- DocumentTermMatrix(tweet_corpus)
tweet_dtm_matrix <- as.matrix(tweet_dtm)

# Examine the Document Term Matrix
str(tweet_dtm_matrix)
```

Appropriately, we get a Document Term Matrix with 1000 rows (representing the 1000 tweets that were queried) and 2441 columns (representing 2441 unique terms that were found in the dataset).

```{r tidy_data}
# Tidy the Document Term Matrix for tidytext analysis
tidy_tweets <- tidy(tweet_dtm)

# Examine the tidied dataset
head(tidy_tweets)
```

The second stage of preprocessing gives us a "tidy dataset" i.e. a data structure that abides by "tidyverse" conventions. By having "term" as a column, this gives us the opportunity to join a sentiment lexicon in the next stage to draw insights about the sentiment of the dataset.

### 6. Sentiment Analysis with tidytext Lexicons

```{r def_bing_polarity}
# Store the "Bing" lexicon from tidytext
bing <- get_sentiments(lexicon = "bing")

# Inner join the Bing lexicon to the Document Term Matrix and generate a Polarity Score
tweet_polarity_bing <-
  tidy_tweets %>% 
  inner_join(bing, by = c("term" = "word")) %>% 
  mutate(index = as.numeric(document)) %>% 
  count(sentiment, index) %>% 
  spread(sentiment, n, fill = 0) %>% 
  mutate(polarity = positive - negative)

# Examine the mutated dataset
str(tweet_polarity_bing)
```

```{r plot_bing}
# Plot Polarity Scores against Index
ggplot(data = tweet_polarity_bing, aes(rev(index), polarity)) +
  geom_smooth() +
  labs(
    title = "Polarity of #CrudeOil tweets (With Bing Lexicon)",
    caption = "Source: Twitter API, Bing Lexicon",
    x = "Time",
    y = "Polarity"
  ) +
  theme_minimal()
```


```{r def_afinn_polarity}
# Store the "AFINN" lexicon from tidytext
afinn <- get_sentiments(lexicon = "afinn")

# Inner join the AFINN lexicon to the Document Term Matrix and generate a Polarity Score
tweet_polarity_afinn <-
  tidy_tweets %>% 
  inner_join(afinn, by = c("term" = "word")) %>% 
  mutate(index = as.numeric(document)) %>%
  count(value, index) %>% 
  group_by(index) %>% 
  summarize(polarity = sum(value * n))

# Examine the mutated dataset
head(tweet_polarity_afinn)
```

```{r plot_afinn}
# Plot Polarity Scores against Index
ggplot(data = tweet_polarity_afinn, aes(rev(index), polarity)) +
  geom_smooth() +
  labs(
    title = "Polarity of #CrudeOil tweets (With AFINN Lexicon)",
    caption = "Source: Twitter API, AFINN Lexicon",
    x = "Time",
    y = "Polarity"
  ) +
  theme_minimal()
```


```{r def_loughran_polarity}
# Store the "Loughran-Macdonald" Sentiment Lexicon from tidytext
loughran <- get_sentiments(lexicon = "loughran")

# Inner join the Bing lexicon to the Document Term Matrix and generate a Polarity Score
tweet_polarity_loughran <-
  tidy_tweets %>% 
  inner_join(loughran, by = c("term" = "word")) %>% 
  mutate(index = as.numeric(document)) %>% 
  group_by(sentiment) %>% 
  summarize(count = sum(count))

# Examine the mutated dataset
head(tweet_polarity_loughran)
```

```{r plot_loughran}
# Plot a distribution of sentiments
ggplot(tweet_polarity_loughran, aes(x = reorder(sentiment, desc(count)), y = count, fill = sentiment)) +
  geom_col() +
  labs(
    title = "Polarity of #CrudeOil tweets (With Loughran-Macdonald Lexicon)",
    caption = "Source: Twitter API, Loughran-Macdonald Lexicon",
    x = "Sentiment",
    y = "Count"
  ) +
  theme(legend.position = "none") +
  theme_minimal()
```


### 7. Comparison Clouds, Commonality Clouds and Pyramind Plots

In this final section, we split the tweet data into Positive and Negative baskets and draw out the most frequent terms per category.

To do so, we structure the data as a Term Document Matrix instead of a Document Term Matrix as a we did before.

```{r segregate_sentiment}
# Extract all "positive-sentiment" tweets and collapse into a single string
pos_terms <-
  crude_df %>% 
  mutate(polarity = polarity(text)$all$polarity) %>% 
  filter(polarity > 0) %>% 
  pull(text) %>% 
  paste(collapse = " ")

# View the joined string
substr(pos_terms, 1, 100)
```

```{r get_negs}
# Extract all "negative-sentiment" tweets and collapse into a single string
neg_terms <-
  crude_df %>% 
  mutate(polarity = polarity(text)$all$polarity) %>% 
  filter(polarity < 0) %>% 
  pull(text) %>% 
  paste(collapse = " ")

# View the joined string
substr(neg_terms, 1, 100)
```

```{r combine_two}
# Concatenate the two strings into a vector with 2 elements
all_terms <- c(pos_terms, neg_terms)

# Convert the vector into a Corpus
corpus_all_terms <- VCorpus(VectorSource(all_terms))

# Convert the Corpus into a TermDocumentMatrix
tdm_all_terms <- TermDocumentMatrix(
  corpus_all_terms,
  control = list(
    weighting = weightTf,
    removePunctuation = TRUE
  )
)

# Convert the TDM into an R matrix object
matrix_all_terms <- as.matrix(tdm_all_terms)
colnames(matrix_all_terms) <- c("Positive", "Negative")

# View the TDM Matrix
tail(matrix_all_terms)
```

```{r plot_comparison, warning=FALSE}
# Plot a Comparison Cloud with the tidied TDM
comparison.cloud(matrix_all_terms, colors = c("green", "red"))
```

```{r plot_commonality, warning=FALSE}
commonality.cloud(matrix_all_terms, colors = "steelblue1")
```

```{r def_comparison}
# Create a data.frame containing 25 of the most common terms to compare
top15_terms <-
  matrix_all_terms %>% 
  as_tibble(rownames = "word") %>% 
  filter_all(all_vars(. > 0)) %>% 
  mutate(difference = Positive - Negative) %>% 
  top_n(15, wt = difference) %>% 
  arrange(desc(difference))

# Examine the top 25 terms
head(top15_terms)
```

```{r plot_pyramid}
# Create a Pyramid Plot to visualize the differences
library(plotrix)

pyramid.plot(
  top15_terms$Positive,
  top15_terms$Negative,
  labels = top15_terms$word,
  top.labels = c("Positive", "Word", "Negative"),
  main = "Words in Common",
  unit = NULL,
  gap = 10,
  space = 0.2
)

```


### 8. Word Networks with qdap

A final form of analysis can tell us what words are most associated with a pre-defined word. We do this with the word_associate() function in qdap which can generate network graphs and/or word clouds around this form of analysis.

```{r plot_network}
# Create a Word Network plot with qdap'
word_associate(
  crude_df$text, 
  match.string = c("trump"), 
  stopwords = c("crude", stopwords("en")), 
  wordcloud = TRUE, 
  cloud.colors = c("gray85", "darkred"),
  nw.label.proportional = TRUE
  )

title(main = "Words Associated with Trump in #CrudeOil Tweets")
```

```{r plot_network2}
# Create a Word Network plot with qdap'
word_associate(
  crude_df$text, 
  match.string = c("fed"), 
  stopwords = c("crude", stopwords("en")), 
  wordcloud = TRUE, 
  cloud.colors = c("gray85", "darkred"),
  nw.label.proportional = TRUE
  )

title(main = "Words Associated with Fed in #CrudeOil Tweets")
```

### 9. Final Conclusions

This project attempted to leverage on key NLP libraries in R to derive new insights about market sentiment in the Crude Oil market.

By using a variety of Sentiment Lexicons, we found that, on average, sentiment for tweets tagged with #crudeoil tended towards the negative. This was unsurprising considering geopolitical developments around the U.S. and Iran, and the effect that those events have on the supply of crude oil.

Interestingly, we note that *price action* for Crude Oil in this timeframe was *overwhelmingly bullish*. (Note only was there the looming threat of war cutting off crude supplies, but the Fed's dovish outlook also provided support for prices). This suggests that a naive interpretation of using negative sentiment to take bearish positions is overwhelmingly simplistic.

Future projects could involve the transformation of such sentiment data into an algorithmic trading strategy with packages such as 'quanstrat'.

```{r export_to_script}
# Export the analysis to an R Script
purl("crude-oil-sentiment-analysis.Rmd")
```

